---
title: "Random Forest PDP/ICE Workflow"
author: "David Leydet"
date: "2024-09-08"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: false
    theme: bootstrap
    df_print: paged
   
---

# **Introduction**

This document contains a draft workflow to building random forests and PD/ICE plots. The workflow includes examples to color the ICE plots by variables.


```{r Set Working Directory, echo=FALSE}

setwd("~/Desktop/University of Utah PhD /Research/r_code/")

```



# **Stack Overflow Examples**

The Ames Housing dataset is from De Cock (2011). It has 82 fields were recorded for 2,930 properties in Ames IA. This version is copies from the AmesHousing package but does not include a few quality columns that appear to be outcomes rather than predictors.

De Cock, D. (2011). "Ames, Iowa: Alternative to the Boston Housing Data as an End of Semester Regression Project," Journal of Statistics Education, Volume 19, Number 3.

```{r}
library(tidymodels)
library(DALEXtra) # DALEX is an R package with a set of tools that help to provide Descriptive mAchine Learning EXplanations ranging from global to local interpretability methods.

# https://uc-r.github.io/dalex


data(ames)

#transmute is an old dplyr function. Mutate is now used.
#create a data frame with sale price, gr liv area, year built, and bldg type
ames_train <- ames %>%
    transmute(Sale_Price = log10(Sale_Price),
              Gr_Liv_Area = as.numeric(Gr_Liv_Area), 
              Year_Built, Bldg_Type)

```


```{r}
#rf model setup using parsnip package

rf_model <- 
    rand_forest(trees = 1000) %>% 
    set_engine("ranger") %>% 
    set_mode("regression")

```


```{r}
#workflow framework to run the random forest

rf_wflow <- 
    workflow() %>% 
    add_formula(
        Sale_Price ~ Gr_Liv_Area + Year_Built + Bldg_Type) %>% 
    add_model(rf_model) 

```

```{r}

## fit the data from our data frame of interest (in this case the ames_train data) to our workflow and model
rf_fit <- rf_wflow %>% fit(data = ames_train)


# explainer - unified representation of the model which can be used with various "explainers"
explainer_rf <- explain_tidymodels(
    rf_fit, 
    data = dplyr::select(ames_train, -Sale_Price), 
    y = ames_train$Sale_Price,
    label = "random forest"
)



```


```{r}

pdp_rf <- model_profile(explainer_rf, N = NULL, 
                        variables = "Gr_Liv_Area", groups = "Bldg_Type")

```


```{r}

as_tibble(pdp_rf$agr_profiles) %>%
    mutate(`_label_` = stringr::str_remove(`_label_`, "random forest_")) %>%
    ggplot(aes(`_x_`, `_yhat_`, color = `_label_`)) +
    geom_line(size = 1.2, alpha = 0.8) +
    labs(x = "Gross living area", 
         y = "Sale Price (log)", 
         color = NULL,
         title = "Partial dependence profile for Ames housing sales",
         subtitle = "Predictions from a random forest model")


```


# **Conflict Data Experiment**

Attempt one to recreate the pdp plots in our HMA random forest by country.


```{r, message=FALSE}
library(tidymodels)
library(DALEXtra) # DALEX is an R package with a set of tools that help to provide Descriptive mAchine Learning EXplanations ranging from global to local interpretability methods.

# https://uc-r.github.io/dalex

## read data in
library(tidyverse)

# load cleaned data
adm.imp = read.csv("../cleaned_data/hma_adm1_imp_df_clean_20240418.csv")


# remove row number variable (X) from df
adm.imp = adm.imp %>% 
  select(-X)

# factor country and adm 1
adm.imp$country = as.factor(adm.imp$country)
adm.imp$adm1 = as.factor(adm.imp$adm1)

# check
str(adm.imp)


```


```{r}
#rf model setup using parsnip package

rf_model <- 
    rand_forest(trees = 1000) %>% 
    set_engine("ranger") %>% 
    set_mode("regression")

```


```{r}
#workflow framework to run the random forest

rf_wflow <- 
    workflow() %>% 
    add_formula(
        con ~ .) %>% 
    add_model(rf_model) 

```

```{r}

## fit the data from our data frame of interest (in this case the ames_train data) to our workflow and model
rf_fit <- rf_wflow %>% fit(data = adm.imp)

```


```{r}

# explainer - unified representation of the model which can be used with various "explainers"
explainer_rf <- explain_tidymodels(
    rf_fit, 
    data = dplyr::select(adm.imp, -con), # data frame without the target variable
    y = adm.imp$con, # target variable
    label = "random forest"
)



```


```{r Pop PDP Build 1}

pdp_rf <- model_profile(explainer_rf, N = NULL, 
                        variables = "pop", groups = "country")

# explainer - I think you can only select one variable to look at pdp wise, then group it by a categorical variable, in this case group by country.

```


```{r Pop PDP Build Label Code Version}



pop.pdp = as_tibble(pdp_rf$agr_profiles) %>%
    mutate(`_label_` = stringr::str_remove(`_label_`, "random forest_"))

pop.pdp$label_1 <- NA

pop.pdp$label_1 <- as.character(pop.pdp$label_1)

for(i in 1:length(unique(pop.pdp$`_label_`))) {
  
  pop.pdp[which(pop.pdp$`_label_` == 
                    unique(pop.pdp$`_label_`)[i]), "label_1"][100, ] <- unique(pop.pdp$`_label_`)[i]
}

  
```

```{r Pop PDP Build Label Code Version}

library(ggrepel)

pop.plot = ggplot(pop.pdp, aes(`_x_`, `_yhat_`, color = `_label_`)) +
  geom_line(size = 1.2, alpha = 0.8, show.legend = FALSE) +
  labs(x = "Population", 
       y = "Conflict Count", 
       color = NULL,
       title = "Population PDP") +
  geom_label_repel(aes(label = label_1), show.legend = FALSE)

pop.plot

```


```{r PDP Code from Brenna}
# code for labeling!

#pop.pdp_1 = pop.pdp

#pop.pdp_1$label_1 <- NA

#pop.pdp_1$label_1 <- as.character(pop.pdp_1$label_1)

#for(i in 1:length(unique(pop.pdp_1$`_label_`))) {
  
  #pop.pdp_1[which(pop.pdp_1$`_label_` == 
                    #unique(pop.pdp_1$`_label_`)[i]), "label_1"][100, ] <- unique(pop.pdp_1$`_label_`)[i]
#}

#library(ggrepel)

#ggplot(pop.pdp_1, aes(`_x_`, `_yhat_`, color = `_label_`)) +
  #geom_line(size = 1.2, alpha = 0.8, show.legend = FALSE) +
  #labs(x = "Population", 
       #y = "Conflict Count", 
       #color = NULL,
       #title = "Population PDP") +
  #geom_label_repel(aes(label = label_1), show.legend = FALSE)
```


```{r Polstab PDP Build 1}

pdp_rf2 <- model_profile(explainer_rf, N = NULL, 
                        variables = "polstab", groups = "country")

# explainer - I think you can only select one variable to look at pdp wise, then group it by a categorical variable, in this case group by country.

```


```{r Polstab PDP Build 2}
## **ORIGINAL CODE**

#polstab.pdp = as_tibble(pdp_rf2$agr_profiles) %>%
    #mutate(`_label_` = stringr::str_remove(`_label_`, "random forest_")) %>%
    #ggplot(aes(`_x_`, `_yhat_`, color = `_label_`)) +
    #geom_line(size = 1.2, alpha = 0.8) +
    #labs(x = "Political Stability", 
         #y = "Conflict Count", 
         #color = NULL,
         #title = "Political Stability PDP")
         #subtitle = "Predictions from a random forest model")

#polstab.pdp 

```
```{r Polstab PDP Build Label Code Version}



polstab.pdp = as_tibble(pdp_rf2$agr_profiles) %>%
    mutate(`_label_` = stringr::str_remove(`_label_`, "random forest_"))

polstab.pdp$label_1 <- NA

polstab.pdp$label_1 <- as.character(polstab.pdp$label_1)

for(i in 1:length(unique(polstab.pdp$`_label_`))) {
  
  polstab.pdp[which(polstab.pdp$`_label_` == 
                    unique(polstab.pdp$`_label_`)[i]), "label_1"][78, ] <- unique(polstab.pdp$`_label_`)[i]
}

  
```

```{r Polstab PDP Build Label Code Version}

library(ggrepel)

polstab.plot = ggplot(polstab.pdp, aes(`_x_`, `_yhat_`, color = `_label_`)) +
  geom_line(size = 1.2, alpha = 0.8, show.legend = FALSE) +
  labs(x = "Political Stability", 
       y = "Conflict Count", 
       color = NULL,
       title = "Political Stability PDP") +
  geom_label_repel(aes(label = label_1), show.legend = FALSE)

polstab.plot

```

```{r Temperature PDP Build 1}

pdp_rf3 <- model_profile(explainer_rf, N = NULL, 
                        variables = "temp", groups = "country")

# explainer - I think you can only select one variable to look at pdp wise, then group it by a categorical variable, in this case group by country.

```


```{r Temperature PDP Build 2}
# **ORIGINAL CODE**

#temp.pdp = as_tibble(pdp_rf3$agr_profiles) %>%
    #mutate(`_label_` = stringr::str_remove(`_label_`, "random forest_")) %>%
    #ggplot(aes(`_x_`, `_yhat_`, color = `_label_`)) +
    #geom_line(size = 1.2, alpha = 0.8) +
    #labs(x = "Temperature", 
         #y = "Conflict Count", 
         #color = NULL,
         #title = "Temperature PDP")
         #subtitle = "Predictions from a random forest model")

#temp.pdp 

```


```{r Temperature PDP Build Label Code Version}



temp.pdp = as_tibble(pdp_rf3$agr_profiles) %>%
    mutate(`_label_` = stringr::str_remove(`_label_`, "random forest_"))

temp.pdp$label_1 <- NA

temp.pdp$label_1 <- as.character(temp.pdp$label_1)

for(i in 1:length(unique(temp.pdp$`_label_`))) {
  
  temp.pdp[which(temp.pdp$`_label_` == 
                    unique(temp.pdp$`_label_`)[i]), "label_1"][100, ] <- unique(temp.pdp$`_label_`)[i]
}

  
```

```{r Temperature PDP Build Label Code Version}

library(ggrepel)

temp.plot = ggplot(temp.pdp, aes(`_x_`, `_yhat_`, color = `_label_`)) +
  geom_line(size = 1.2, alpha = 0.8, show.legend = FALSE) +
  labs(x = "Temperature", 
       y = "Conflict Count", 
       color = NULL,
       title = "Temperature PDP") +
  geom_label_repel(aes(label = label_1), show.legend = FALSE)

temp.plot

```


```{r Precipitation PDP Build 1}

pdp_rf4 <- model_profile(explainer_rf, N = NULL, 
                        variables = "precip", groups = "country")

# explainer - I think you can only select one variable to look at pdp wise, then group it by a categorical variable, in this case group by country.

```


```{r Precipitation PDP Build 2}
# **ORIGINAL CODE**

#precip.pdp = as_tibble(pdp_rf4$agr_profiles) %>%
    #mutate(`_label_` = stringr::str_remove(`_label_`, "random forest_")) %>%
    #ggplot(aes(`_x_`, `_yhat_`, color = `_label_`)) +
    #geom_line(size = 1.2, alpha = 0.8) +
    #labs(x = "Precipitation", 
         #y = "Conflict Count", 
         #color = NULL,
         #title = "Precipitation PDP")
         #subtitle = "Predictions from a random forest model")

#precip.pdp 

```
```{r Precipitation PDP Build Label Code Version}



precip.pdp = as_tibble(pdp_rf4$agr_profiles) %>%
    mutate(`_label_` = stringr::str_remove(`_label_`, "random forest_"))

precip.pdp$label_1 <- NA

precip.pdp$label_1 <- as.character(precip.pdp$label_1)

for(i in 1:length(unique(precip.pdp$`_label_`))) {
  
  precip.pdp[which(precip.pdp$`_label_` == 
                    unique(precip.pdp$`_label_`)[i]), "label_1"][100, ] <- unique(precip.pdp$`_label_`)[i]
}

  
```

```{r Precipitation PDP Build Label Code Version}

library(ggrepel)

precip.plot = ggplot(precip.pdp, aes(`_x_`, `_yhat_`, color = `_label_`)) +
  geom_line(size = 1.2, alpha = 0.8, show.legend = FALSE) +
  labs(x = "Precipitation", 
       y = "Conflict Count", 
       color = NULL,
       title = "Precipitation PDP") +
  geom_label_repel(aes(label = label_1), show.legend = FALSE)

precip.plot

```



```{r Combined PDP Figure - ADM1 missForest}
library(ggpubr)

ggarrange(pop.plot, temp.plot, polstab.plot, precip.plot, ncol = 2, nrow = 2)#, common.legend = TRUE, legend = "right")

#gridExtra::grid.arrange(pop.pdp, polstab.pdp, nrow = 2)


```


