---
title: "Topic Modeling Example"
author: "David Leydet"
date: "2024-10-02"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: false
    theme: bootstrap
    df_print: paged
   
---

# **Introduction**

The following is example code and discussion from Schweinberger, Martin. 2024. Topic Modeling with R. Brisbane: The University of Queensland. url: https://slcladal.github.io/topicmodels.html (Version 2024.04.22). There code and explanations can be found [here](https://ladal.edu.au/topicmodels.html).




# **Setup**

```{r}

setwd("../r_code/")

```


```{r}
# install the required pacakges

# install packages
#install.packages("tm")
#install.packages("topicmodels")
#install.packages("reshape2")
#install.packages("ggplot2")
#install.packages("wordcloud")
#install.packages("pals")
#install.packages("SnowballC")
#install.packages("lda")
#install.packages("ldatuning")
#install.packages("kableExtra")
#install.packages("DT")
#install.packages("flextable")

# install klippy for copy-to-clipboard button in code chunks
#install.packages("remotes")
#remotes::install_github("rlesur/klippy")
```



```{r, message = FALSE}
# load libraries

# set options
options(stringsAsFactors = F)         # no automatic data transformation
options("scipen" = 100, "digits" = 4) # suppress math annotation

# load packages
library(knitr) 
library(kableExtra) 
library(DT)
library(tm)
library(topicmodels)
library(reshape2)
library(ggplot2)
library(wordcloud)
library(pals)
library(SnowballC)
library(lda)
library(ldatuning)
library(flextable)

# activate klippy for copy-to-clipboard button
klippy::klippy()
```



# **Topic Modeling Intro**

The process starts as usual with the reading of the corpus data. For this tutorial we will analyze State of the Union Addresses (SOTU) by US presidents and investigate how the topics that were addressed in the SOTU speeches changeover time. The 231 SOTU addresses are rather long documents. Documents lengths clearly affects the results of topic modeling. For very short texts (e.g. Twitter posts) or very long texts (e.g. books), it can make sense to concatenate/split single documents to receive longer/shorter textual units for modeling.

For the SOTU speeches for instance, we infer the model based on paragraphs instead of entire speeches. By manual inspection / qualitative inspection of the results you can check if this procedure yields better (interpretable) topics. In sotu_paragraphs.csv, we provide a paragraph separated version of the speeches.

For text preprocessing, we remove stopwords, since they tend to occur as “noise” in the estimated topics of the LDA model.


```{r}

# load data
textdata <- base::readRDS(url("https://slcladal.github.io/data/sotu_paragraphs.rda", "rb"))

# load stopwords
english_stopwords <- readLines("https://slcladal.github.io/resources/stopwords_en.txt", encoding = "UTF-8")

# create corpus object
corpus <- Corpus(DataframeSource(textdata))

# Preprocessing chain
processedCorpus <- tm_map(corpus, content_transformer(tolower))
processedCorpus <- tm_map(processedCorpus, removeWords, english_stopwords)
processedCorpus <- tm_map(processedCorpus, removePunctuation, preserve_intra_word_dashes = TRUE)
processedCorpus <- tm_map(processedCorpus, removeNumbers)
processedCorpus <- tm_map(processedCorpus, stemDocument, language = "en")
processedCorpus <- tm_map(processedCorpus, stripWhitespace)

```


## **Model Calculation**


After the preprocessing, we have two corpus objects: processedCorpus, on which we calculate an LDA topic model (Blei, Ng, and Jordan 2003b). To this end, stopwords, i.e. function words that have relational rather than content meaning, were removed, words were stemmed and converted to lowercase letters and special characters were removed. The second corpus object corpus serves to be able to view the original texts and thus to facilitate a qualitative control of the topic model results.

We now calculate a topic model on the processedCorpus. For this purpose, a DTM of the corpus is created. In this case, we only want to consider terms that occur with a certain minimum frequency in the body. This is primarily used to speed up the model calculation.


```{r}

# compute document term matrix (DTM) with terms >= minimumFrequency
minimumFrequency <- 5

# build the DTM
DTM <- DocumentTermMatrix(processedCorpus, control = list(bounds = list(global = c(minimumFrequency, Inf))))


# have a look at the number of documents and terms in the matrix
dim(DTM)

```



```{r}

# due to vocabulary pruning, we have empty rows in our DTM ***(how do we know this?)***
# LDA does not like this. So we remove those docs from the
# DTM and the metadata

# create an ID for rows that are not empty (>0)
sel_idx <- slam::row_sums(DTM) > 0

# select those rows for the new DTM
DTM <- DTM[sel_idx, ]

# select those rows for the original textdata
textdata <- textdata[sel_idx, ]

```




As an unsupervised machine learning method, topic models are suitable for the exploration of data. The calculation of topic models aims to determine the proportionate composition of a fixed number of topics in the documents of a collection. It is useful to experiment with different parameters in order to find the most suitable parameters for your own analysis needs.

For parameterized models such as Latent Dirichlet Allocation (LDA), the number of topics K is the most important parameter to define in advance. How an optimal K should be selected depends on various factors. If K is too small, the collection is divided into a few very general semantic contexts. If K is too large, the collection is divided into too many topics of which some may overlap and others are hardly interpretable.

An alternative to deciding on a set number of topics is to extract parameters form a models using a rage of number of topics. This approach can be useful when the number of topics is not theoretically motivated or based on closer, qualitative inspection of the data. In the example below, the determination of the optimal number of topics follows Murzintcev (n.d.), but we only use two metrics (CaoJuan2009 and Deveaud2014) - it is highly recommendable to inspect the results of the four metrics available for the FindTopicsNumber function (Griffiths2004, CaoJuan2009, Arun2010, and Deveaud2014).


```{r}

# create models with different number of topics
result <- ldatuning::FindTopicsNumber(
  DTM,
  topics = seq(from = 2, to = 20, by = 1),
  metrics = c("CaoJuan2009",  "Deveaud2014"),
  method = "Gibbs",
  control = list(seed = 77),
  verbose = TRUE
)

```


We can now plot the results. In this case, we have only use two methods CaoJuan2009 and Griffith2004. The best number of topics shows low values for CaoJuan2009 and high values for Griffith2004 (optimally, several methods should converge and show peaks and dips respectively for a certain number of topics).


```{r}

# visualize
FindTopicsNumber_plot(result)


# do we select where it converges? Is 20 the "optimal" number of topics?

```


For our first analysis, however, we choose a thematic “resolution” of K = 20 topics. In contrast to a resolution of 100 or more, this number of topics can be evaluated qualitatively very easy.



```{r}

# number of topics
K <- 20

# set random number generator seed
set.seed(9161)

# compute the LDA model, inference via 1000 (500?) iterations of Gibbs sampling
topicModel <- LDA(DTM, K, method="Gibbs", control=list(iter = 500, verbose = 25))

```


Depending on the size of the vocabulary, the collection size and the number K, the inference of topic models can take a very long time. This calculation may take several minutes. If it takes too long, reduce the vocabulary in the DTM by increasing the minimum frequency in the previous step.

The topic model inference results in two (approximate) posterior probability distributions: a distribution theta over K topics within each document and a distribution beta over V terms within each topic, where V represents the length of the vocabulary of the collection (V = 4278). Let’s take a closer look at these results:

```{r}

# have a look a some of the results (posterior distributions)
tmResult <- posterior(topicModel)

# format of the resulting object
attributes(tmResult)

```



```{r}

nTerms(DTM)              # lengthOfVocab (terms)

```



```{r}

# topics are probability distributions over the entire vocabulary
beta <- tmResult$terms   # get beta from results

dim(beta)                # K distributions over nTerms(DTM) terms

```




```{r}

rowSums(beta)            # rows in beta sum to 1

```




```{r}

nDocs(DTM)               # size of collection (documents)

```



```{r}

# for every document we have a probability distribution of its contained topics
theta <- tmResult$topics 
dim(theta)               # nDocs(DTM) distributions over K topics

```




```{r}

rowSums(theta)[1:10]     # rows in theta sum to 1

```



Let’s take a look at the 10 most likely terms within the term probabilities beta of the inferred topics(only the first 8 are shown below).



```{r}

terms(topicModel, 10)

```


```{r}

exampleTermData <- terms(topicModel, 10)
exampleTermData[, 1:8]


```


For the next steps, we want to give the topics more descriptive names than just numbers. Therefore, we simply concatenate the five most likely terms of each topic to a string that represents a pseudo-name for each topic.


```{r}

top5termsPerTopic <- terms(topicModel, 5)
topicNames <- apply(top5termsPerTopic, 2, paste, collapse=" ")


```


## **Visualization of Words and Topics**

Although wordclouds may not be optimal for scientific purposes they can provide a quick visual overview of a set of terms. Let’s look at some topics as wordcloud.

In the following code, you can change the variable topicToViz with values between 1 and 20 to display other topics.


```{r}

# visualize topics as word cloud
topicToViz <- 11 # change for your own topic of interest (what purpose does this line serve?)

topicToViz <- grep('mexico', topicNames)[1] # Or select a topic by a term contained in its name

# select to 40 most probable terms from the topic by sorting the term-topic-probability vector in decreasing order

top40terms <- sort(tmResult$terms[topicToViz,], decreasing=TRUE)[1:40]
words <- names(top40terms)

# extract the probabilites of each of the 40 terms
probabilities <- sort(tmResult$terms[topicToViz,], decreasing=TRUE)[1:40]

# visualize the terms as wordcloud
mycolors <- brewer.pal(8, "Dark2")
wordcloud(words, probabilities, random.order = FALSE, color = mycolors)


```



Let us now look more closely at the distribution of topics within individual documents. To this end, we visualize the distribution in 3 sample documents.

Let us first take a look at the contents of three sample documents:


```{r}

exampleIds <- c(2, 100, 200)

lapply(corpus[exampleIds], as.character)


```



```{r}

exampleIds <- c(2, 100, 200)
print(paste0(exampleIds[1], ": ", substr(content(corpus[[exampleIds[1]]]), 0, 400), '...'))

```


```{r}

exampleIds <- c(2, 100, 200)
print(paste0(exampleIds[2], ": ", substr(content(corpus[[exampleIds[1]]]), 0, 400), '...'))

```

```{r}

exampleIds <- c(2, 100, 200)
print(paste0(exampleIds[3], ": ", substr(content(corpus[[exampleIds[1]]]), 0, 400), '...'))

```


After looking into the documents, we visualize the topic distributions within the documents.



```{r}

N <- length(exampleIds)

# get topic proportions form example documents
topicProportionExamples <- theta[exampleIds,] # retrieve the proportions (probabilities?)

colnames(topicProportionExamples) <- topicNames # change the column names to the topic names (of which there are 20)

vizDataFrame <- melt(cbind(data.frame(topicProportionExamples), document = factor(1:N)), variable.name = "topic", id.vars = "document")  # build the dataframe

#Visualize
ggplot(data = vizDataFrame, aes(topic, value, fill = document), ylab = "proportion") + 
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  
  coord_flip() +
  facet_wrap(~ document, ncol = N)


```


## **Topic Distributions**


The figure above shows how topics within a document are distributed according to the model. In the current model all three documents show at least a small percentage of each topic. However, two to three topics dominate each document.

The topic distribution within a document can be controlled with the Alpha-parameter of the model. Higher alpha priors for topics result in an even distribution of topics within a document. Low alpha priors ensure that the inference process distributes the probability mass on a few topics for each document.

In the previous model calculation the alpha-prior was automatically estimated in order to fit to the data (highest overall probability of the model). However, this automatic estimate does not necessarily correspond to the results that one would like to have as an analyst. Depending on our analysis interest, we might be interested in a more peaky/more even distribution of topics in the model.

Now let us change the alpha prior to a lower value to see how this affects the topic distributions in the model.




```{r}

# see alpha from previous model
attr(topicModel, "alpha") 

```



```{r}
# run another model
# set the alpha parameter lower - to 0.2
topicModel2 <- LDA(DTM, K, method="Gibbs", control=list(iter = 500, verbose = 25, alpha = 0.2))

```



```{r}

tmResult <- posterior(topicModel2)
theta <- tmResult$topics
beta <- tmResult$terms
topicNames <- apply(terms(topicModel2, 5), 2, paste, collapse = " ")  # reset topicnames

```

Now visualize the topic distributions in the three documents again. What are the differences in the distribution structure?


```{r}

# get topic proportions form example documents
topicProportionExamples <- theta[exampleIds,]
colnames(topicProportionExamples) <- topicNames
vizDataFrame <- melt(cbind(data.frame(topicProportionExamples), document = factor(1:N)), variable.name = "topic", id.vars = "document")  

# visualize
ggplot(data = vizDataFrame, aes(topic, value, fill = document), ylab = "proportion") + 
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  
  coord_flip() +
  facet_wrap(~ document, ncol = N)

```


## **Topic Ranking**

First, we try to get a more meaningful order of top terms per topic by re-ranking them with a specific score (Chang et al. 2009). The idea of re-ranking terms is similar to the idea of TF-IDF. The more a term appears in top levels w.r.t. its probability, the less meaningful it is to describe the topic. Hence, the scoring advanced favors terms to describe a topic.

```{r}
# re-rank top topic terms for topic names
topicNames <- apply(lda::top.topic.words(beta, 5, by.score = T), 2, paste, collapse = " ")

```


### **Approach 1**

We sort topics according to their probability within the entire collection:

```{r}

# What are the most probable topics in the entire collection?

topicProportions <- colSums(theta) / nDocs(DTM)  # mean probabilities over all paragraphs
names(topicProportions) <- topicNames     # assign the topic names we created before
sort(topicProportions, decreasing = TRUE) # show summed proportions in decreased order

```


```{r}

# sort the probabilities in descending order

soP <- sort(topicProportions, decreasing = TRUE)
paste(round(soP, 5), ":", names(soP))

```

We recognize some topics that are way more likely to occur in the corpus than others. These describe rather general thematic coherence. Other topics correspond more to specific contents.



### **Approach 2**

We count how often a topic appears as a primary topic within a paragraph This method is also called Rank-1.

```{r}

countsOfPrimaryTopics <- rep(0, K)

names(countsOfPrimaryTopics) <- topicNames

for (i in 1:nDocs(DTM)) {
  topicsPerDoc <- theta[i, ] # select topic distribution for document i
  # get first element position from ordered list
  primaryTopic <- order(topicsPerDoc, decreasing = TRUE)[1] 
  countsOfPrimaryTopics[primaryTopic] <- countsOfPrimaryTopics[primaryTopic] + 1
}

sort(countsOfPrimaryTopics, decreasing = TRUE)

```


```{r}

so <- sort(countsOfPrimaryTopics, decreasing = TRUE)
paste(so, ":", names(so))

```

We see that sorting topics by the Rank-1 method places topics with rather specific thematic coherences in upper ranks of the list.

This sorting of topics can be used for further analysis steps such as the semantic interpretation of topics found in the collection, the analysis of time series of the most important topics or the filtering of the original collection based on specific sub-topics.

## **Filtering Documents**

The fact that a topic model conveys of topic probabilities for each document, resp. paragraph in our case, makes it possible to use it for thematic filtering of a collection. AS filter we select only those documents which exceed a certain threshold of their probability value for certain topics (for example, each document which contains topic X to more than 20 percent).

In the following, we will select documents based on their topic content and display the resulting document quantity over time.




```{r}

topicToFilter <- 6  # you can set this manually ...
# ... or have it selected by a term in the topic name (e.g. 'children')

topicToFilter <- grep('children', topicNames)[1] 

topicThreshold <- 0.2 # more than 20%
selectedDocumentIndexes <- which(theta[, topicToFilter] >= topicThreshold)
filteredCorpus <- corpus[selectedDocumentIndexes]

# show length of filtered corpus
filteredCorpus

```

Our filtered corpus contains 0 documents related to the topic NA to at least 20 %.


## **Topic Proportions over Time**

In a last step, we provide a distant view on the topics in the data over time. For this, we aggregate mean topic proportions per decade of all SOTU speeches. These aggregated topic proportions can then be visualized, e.g. as a bar plot.

```{r}

# append decade information for aggregation
textdata$decade <- paste0(substr(textdata$date, 0, 3), "0")

# get mean topic proportions per decade
topic_proportion_per_decade <- aggregate(theta, by = list(decade = textdata$decade), mean)

# set topic names to aggregated columns
colnames(topic_proportion_per_decade)[2:(K+1)] <- topicNames

# reshape data frame
vizDataFrame <- melt(topic_proportion_per_decade, id.vars = "decade")

# plot topic proportions per decade as bar plot
ggplot(vizDataFrame, aes(x=decade, y=value, fill=variable)) + 
  geom_bar(stat = "identity") + ylab("proportion") + 
  scale_fill_manual(values = paste0(alphabet(20), "FF"), name = "decade") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

The visualization shows that topics around the relation between the federal government and the states as well as inner conflicts clearly dominate the first decades. Security issues and the economy are the most important topics of recent SOTU addresses.



# **Citation**

Schweinberger, Martin. 2024. Topic Modeling with R. Brisbane: The University of Queensland. url: https://slcladal.github.io/topicmodels.html (Version 2024.04.22).




References
Blei, David M., Andrew Y. Ng, and Michael I. Jordan. 2003b. “Latent Dirichlet Allocation.” Journal of Machine Learning Research 3 (3): 993–1022.
———. 2003a. “Latent Dirichlet Allocation.” Journal of Machine Learning Research 3: 993–1022.
Brookes, Gavin, and Tony McEnery. 2019. “The Utility of Topic Modelling for Discourse Studies.” Discourse Studies 21 (1): 3–21. https://doi.org/10.1177/14614456188140.
Busso, Luciana, Monika Petyko, Steven Atkins, and Tim Grant. 2022. “Operation Heron: Latent Topic Changes in an Abusive Letter Series.” Corpora 17 (2): 225–58.
Chang, Jonathan, Sean Gerrish, Chong Wang, Jordan L. Boyd-graber, and David M. Blei. 2009. “Reading Tea Leaves: How Humans Interpret Topic Models.” In Advances in Neural Information Processing Systems 22, edited by Yoshua Bengio, Dale Schuurmans, John D. Lafferty, Christopher K. Williams, and Aron Culotta, 288–96. Curran. http://papers.nips.cc/paper/3700-reading-tea-leaves-how-humans-interpret-topic-models.pdf.
Gerlach, Martin, Tiago P. Peixoto, and Eduardo G. Altmann. 2018. “A Network Approach to Topic Models.” Science Advances 4: eaar1360.
Gillings, Mark, and Andrew Hardie. 2022. “The Interpretation of Topic Models for Scholarly Analysis: An Evaluation and Critique of Current Practice.” Digital Scholarship in the Humanities. https://doi.org/10.1093/llc/fqac075.
Hyland, Conor C., Yang Tao, Lida Azizi, Martin Gerlach, Tiago P. Peixoto, and Eduardo G. Altmann. 2021. “Multilayer Networks for Text Analysis with Multiple Data Types.” EPJ Data Science 10: 33.
Murzintcev, Nikita. n.d. “Select Number of Topics for LDA Model.” https://cran.r-project.org/web/packages/ldatuning/vignettes/topics.html.
Roberts, Margaret E., Brandon M. Stewart, and Dustin Tingley. 2016. “Navigating the Local Modes of Big Data: The Case of Topic Models.” In Computational Social Science: Discovery and Prediction, edited by R. Michael Alvarez, 51–97. Cambridge University Press.
Silge, Julia, and David Robinson. 2017. Text Mining with r: A Tidy Approach. O’Reilly.
Wiedemann, Gregor, and Andreas Niekler. 2017. “Hands-on: A Five Day Text Mining Course for Humanists and Social Scientists in R.” In Proceedings of the Workshop on Teaching NLP for Digital Humanities (Teach4DH), Berlin, Germany, September 12, 2017., 57–65. http://ceur-ws.org/Vol-1918/wiedemann.pdf.


































